---
title: "Rindalyzer"
author: "Mauro Esposito"
date: "`r Sys.Date()`"
output: html_document
params:
    results: ""
    fragments: ""
    fastq_r1: ""
    fastq_r2: ""
    reference: ""
    bed: ""
    mismatchPercentage: ""
    genome: ""
    reportDir: ""
---

```{r echo=FALSE, message=FALSE}
# Load packages
library(ggplot2)
library(tidyr)
library(dplyr)
library(magrittr)
library(here)

# Load custom functions
source(paste0(here(), "/Filters_Functions.R"))

# Update the log file
logUpdater(modality = "Update", 
           message = "\nRunning the Report module", 
           outDir = params$reportDir)
```

# Input

Rindalyzer was run by using:

-   FASTQ files, `r paste(params$fastq_r1, "and", params$fastq_r2)`
-   Consensus file, `r params$reference`
-   Genome file, `r params$genome`
-   BED file, `r params$bed`
-   Percentage of allowed mismatches, `r params$mismatchPercentage`

# Preliminary analyses

```{r echo=FALSE}
# Open the table with the results
tableResults <- read.delim(file = params$results, header = T, sep = "\t", stringsAsFactors = F)

# Re-arrange the table
tableFragments <- tableResults[,2:6]
colnames(tableFragments)[1] <- "Fragment_type"
tableFragments %<>%
    gather(ConsensusPreFilter, ConsensusPostFilter, GenomePreFilter, GenomePostFilter, 
           key = "Filtering_step", value = "nFragments")
tableFragments %<>%
    group_by(Fragment_type) %>%
    mutate(Percentage = nFragments/max(nFragments)*100) %>%
    ungroup()
tableFragments <- as.data.frame(tableFragments)
tableFragments[,"Filtering_step"]=factor(x = tableFragments[,"Filtering_step"], 
                                         levels = c("ConsensusPreFilter", "ConsensusPostFilter", 
                                                    "GenomePreFilter", "GenomePostFilter"))

# Counting
insPre <- tableFragments[tableFragments$Fragment_type=="Inside" & tableFragments$Filtering_step=="ConsensusPreFilter",
                         "nFragments"]
outPre <- tableFragments[tableFragments$Fragment_type=="Outside" & tableFragments$Filtering_step=="ConsensusPreFilter",
                         "nFragments"]
insPost <- tableFragments[tableFragments$Fragment_type=="Inside" & tableFragments$Filtering_step=="GenomePostFilter",
                         "nFragments"]
outPost <- tableFragments[tableFragments$Fragment_type=="Outside" & tableFragments$Filtering_step=="GenomePostFilter",
                         "nFragments"]
```

The starting number of Inside and Outside fragments is `r paste(insPre, "and", outPre)`, respectively. After the application of the filters, the number of Inside and Outside fragments is instead `r paste(insPost, "and", outPost)`, respectively.

```{r echo=FALSE}
# Update the log file
logUpdater(modality = "Update", 
           message = "1. Producing the Figure 1", 
           outDir = params$reportDir)

# Plot
ggplot(data = tableFragments, mapping = aes(x = Filtering_step, y = nFragments, 
                                            group = Fragment_type, color = Fragment_type)) +
    geom_line() +
    geom_point() +
    labs(x = "Steps", y = "Number of fragments") +
    guides(color=guide_legend(title="Type of fragment"))
```

After the application of the filters, the number of Inside fragments decreases of `r round(100-((insPost*100)/insPre), 2)`% while the number of Outside fragments decreases of `r round(100-((outPost*100)/outPre), 2)`%

```{r echo=FALSE}
# Update the log file
logUpdater(modality = "Update", 
           message = "2. Producing the Figure 2", 
           outDir = params$reportDir)

# Plot
ggplot(data = tableFragments, mapping = aes(x = Filtering_step, y = Percentage, 
                                            group = Fragment_type, color = Fragment_type)) +
    geom_line() +
    geom_point() +
    labs(x = "Steps", y = "Percentage of fragments")
```

```{r echo=FALSE}
# Compute the ratio
ratio <- (tableResults[1,6]+1)/(tableResults[2,6]+1)

# Re-arrange the table
tableRatio <- (tableResults[1, 3:6]+1)/(tableResults[2, 3:6]+1)
tableRatio %<>%
    gather(ConsensusPreFilter, ConsensusPostFilter, GenomePreFilter, GenomePostFilter, 
           key = "Filtering_step", value = "Ratio")
tableRatio[,"Filtering_step"]=factor(x = tableRatio[,"Filtering_step"], 
                                     levels = c("ConsensusPreFilter", "ConsensusPostFilter", 
                                                "GenomePreFilter", "GenomePostFilter"))
```

# Result

The analysis showed a final Inside/Outside ratio of `r ratio`.

```{r echo=FALSE}
# Update the log file
logUpdater(modality = "Update", 
           message = "3. Producing the Figure 3", 
           outDir = params$reportDir)

ggplot(data = tableRatio, mapping = aes(x = Filtering_step, y = Ratio)) +
    geom_line(group = 1) +
    geom_point() +
    labs(x = "Steps", y = "Inside/Outside")
```

# Coverage analysis

Study of the Inside/Outside fragments distribution along the consensus sequence.

```{r echo=FALSE, message=FALSE}
# Open the table with the info for each read
tableConsensus <- read.delim(file = params$fragments, header = T, sep = "\t", stringsAsFactors = F)

# Compute the length of the consensus length
consensusLength <- consLen(consRef = params$reference)

for(col in colnames(tableConsensus)[6:9]){
    # Calculate the coverage along the consensus sequence
    temp <- tableConsensus[,c("realStart", "realEnd", col)]
    df <- data.frame("Position"=1:consensusLength, "Inside"=0, "Outside"=0)
    for(p in 1:consensusLength){
        df[p, "Inside"] <- sum(temp[,3]=="insIns" & p>=temp$realStart & p<=temp$realEnd)
        df[p, "Outside"] <- sum(temp[,3]=="insOut" & p>=temp$realStart & p<=temp$realEnd)
    }
    df$Inside <- df$Inside/sum(temp[,3]=="insIns")*100
    df$Outside <- df$Outside/sum(temp[,3]=="insOut")*100
    
    # Re-arrange the table
    df %<>%
        gather(Inside, Outside, 
           key = "Fragment_type", value = "Percentage")
    df[,"Step"] <- col
    
    # Store the result
    if(!exists("coverage")){
        coverage <- df
    }else{
        coverage <- rbind(coverage, df)
    }
}

# Re-arrange the table
coverage[,"Step"]=factor(x = coverage[,"Step"], 
                         levels = c("ConsensusPreFilter", "ConsensusPostFilter", 
                                    "GenomePreFilter", "GenomePostFilter"))
quartile <- summary(c(min(coverage$Position), max(coverage$Position)))
coverage[,"Quartile"] <- "to_fill"
coverage[coverage$Position>=quartile[1] & coverage$Position<=quartile[2], "Quartile"] <- "First"
coverage[coverage$Position>=quartile[2] & coverage$Position<=quartile[3], "Quartile"] <- "Second"
coverage[coverage$Position>=quartile[3] & coverage$Position<=quartile[5], "Quartile"] <- "Third"
coverage[coverage$Position>=quartile[5] & coverage$Position<=quartile[6], "Quartile"] <- "Fourth"

temp <- coverage[,2:5] %>%
    group_by(Quartile, Step, Fragment_type) %>%
    summarise("MeanPercentage" = mean(Percentage))
temp <- as.data.frame(temp)

temp[,"Quartile"]=factor(x = temp[,"Quartile"], 
                       levels = c("First", "Second", 
                                  "Third", "Fourth"))

# Update the log file
logUpdater(modality = "Update", 
           message = "4. Producing the Figure 4", 
           outDir = params$reportDir)

# Plot
ggplot(data = temp, mapping = aes(x = Quartile, y = MeanPercentage, fill = Fragment_type)) +
        geom_bar(stat = "identity", position = "dodge", color = "black") +
        guides(fill=guide_legend(title="Type of fragment")) +
        facet_wrap(~Step)

ggplot(data = coverage, mapping = aes(x = Position, y = Percentage, color = Fragment_type)) +
        geom_line() +
        guides(color=guide_legend(title="Type of fragment")) +
        facet_wrap(~Step)
```

# Citation

If you find Rindalyzer useful for your research, please cite:
Esposito, M. et al. Transposons Acting as Competitive Endogenous RNAs: In-Silico Evidence from Datasets Characterised by L1 Overexpression. Biomedicines 2022, 10, 3279. https://doi.org/10.3390/biomedicines10123279